---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions, echo = FALSE} 
library(knitr)
opts_chunk$set(cache = TRUE, echo = TRUE)
# Disable scientific notation
options(scipen=999)
```

``` {r packages, echo = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
```

## Loading and preprocessing the data
First I am loading and pre-processing the data (named *"activity"*).
``` {r Load the data}
# Read in the data
activity <- read.csv(unz("activity.zip", "activity.csv"), sep=",")
# Look at the data and classes
head(activity, n =10)
tail(activity, n =10)
str(activity)
# Transfer the date column into data class
activity$date <- as.Date(activity$date)
# Check
str(activity)
```
## Exploratory Analysis
Look at the summary statistic for the dataset.
``` {r}
summary(activity)

```

## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day.  
``` {r}
FirstDay <- min(activity$date)
LastDay <- max(activity$date)

TotalSteps <- sum(activity$steps, na.rm = TRUE)
TotalSteps

```
The total number of steps within this time period (`r FirstDay` till `r LastDay`) 
is `r TotalSteps`.

The total number of steps per day are:
``` {r}
DailySteps <- tapply(activity$steps, activity$date, sum)
DailySteps

```


2. Make a histogram of the total number of steps taken each day.
``` {r}
g <- ggplot(as.data.frame(DailySteps), aes(DailySteps))+
    geom_histogram(col = "gray", 
                   bins = 8)+
    labs(title = "Histogram of the total number of steps per day",
         x = "Number of steps",
         y = "Frequency")+
    theme_bw()
suppressWarnings(print(g))

```

3. Calculate and report the mean and median of the total number of steps taken per day.  
Mean:
``` {r}
# Average step count for all days
AverageSteps <- mean(DailySteps, na.rm = TRUE)
AverageSteps
```
Median:
``` {r}
# Median step count for all days
MedianSteps <- median(DailySteps, na.rm = TRUE)
MedianSteps

```
The average total step count within this time period (`r FirstDay` till `r LastDay`) 
is `r round(AverageSteps)`. The median is `r round(MedianSteps)`.

## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis).

``` {r}
# Calculate average steps per interval
mean_interval <- tapply(activity$steps, activity$interval, mean, na.rm = T)
# Plot the interval time series plot with ggplot2 function
g2 <- ggplot(as.data.frame(mean_interval), aes(unique(activity$interval), mean_interval))
g2 + geom_line()+
    labs(title = "Time Series Plot of the average steps for each time-interval",
         x = "5-Minute Intervals",
         y = "Average Number of Steps")+
    theme_bw()+
    theme(plot.title = element_text(hjust = 0.5))
```

2. Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?
``` {r}
which.max(mean_interval)
mean_interval[which.max(mean_interval)]


```
The `r which.max(mean_interval)`. Interval namely, the following: `r names(mean_interval[which.max(mean_interval)])` has the
most steps on average (`r round(mean_interval[which.max(mean_interval)], 2)` steps).

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

``` {r}

StepsNA <- sum(is.na(activity$steps))
DateNA <- sum(is.na(activity$date))
IntervalNA <- sum(is.na(activity$interval))

```
There are `r StepsNA` missing values in the "steps" variable, `r DateNA` in 
"dates", and `r IntervalNA` in "intervals".  

2. Devise a strategy for filling in all of the missing values in the dataset. 
The strategy does not need to be sophisticated. For example, you could use the 
mean/median for that day, or the mean for that 5-minute interval, etc.  
  
The missing values in the "steps" variable will be filled with the average 
(mean) of that interval.


3. Create a new dataset that is equal to the original dataset but with the 
missing data filled in.

``` {r}

activityFilled <- activity
for (i in unique(activityFilled$interval)) {
        activityFilled$steps[is.na(activityFilled$steps) & activityFilled$interval == i] <- round(mean(activityFilled$steps[activityFilled$interval == i], na.rm = T), 2)
}

```

4. Make a histogram of the total number of steps taken each day and calculate 
and report the mean and median total number of steps taken per day. Do these 
values differ from the estimates from the first part of the assignment? What 
is the impact of imputing missing data on the estimates of the total daily 
number of steps?

``` {r}

DailyStepsFilled <- tapply(activityFilled$steps, activityFilled$date, sum)
DailyStepsFilled

ggplot(as.data.frame(DailyStepsFilled), aes(DailyStepsFilled))+
    geom_histogram(col = "gray", 
                   bins = 8)+
    labs(title = "Histogram of the total number of steps per day",
         x = "Number of steps",
         y = "Frequency")+
    theme_bw()
```

``` {r}

# Average step count for all days (Filled data)
AverageStepFilled <- mean(DailyStepsFilled)
AverageSteps - AverageStepFilled

# Median step count for all days (Filled data)
MedianStepsFilled <- median(DailyStepsFilled)
MedianSteps - MedianStepsFilled

```
**Conclusion:** The differences between the mean and median values for the daily
steps are really small when using the average interval value as filler for the 
NAs.

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and 
“weekend” indicating whether a given date is a weekday or weekend day.

``` {r}
activityFilled$Weekday <- as.POSIXlt(activityFilled$date)
activityFilled$Weekday <- weekdays(activityFilled$Weekday) == "Samstag" | weekdays(activityFilled$Weekday) == "Sonntag"
activityFilled$Weekday <- factor(activityFilled$Weekday, levels = c(F, T), labels = c("weekday", "weekend"))
table(activityFilled$Weekday)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged 
across all weekday days or weekend days (y-axis).

``` {r}
# Calculate average steps per interval
mean_steps_weekday <- tapply(activityFilled$steps[activityFilled$Weekday == "weekday"], activityFilled$interval[activityFilled$Weekday == "weekday"], mean)

mean_steps_weekend <- tapply(activityFilled$steps[activityFilled$Weekday == "weekend"], activityFilled$interval[activityFilled$Weekday == "weekend"], mean)

# Plot the interval time series plot with base function
par(mfrow = c(2,1))
plot(mean_steps_weekday, 
     type = "l", 
     main = "Time Series Plot of the average steps for each time-interval: Weekdays",
     cex.main = .9,
     xlab = "5-Minute Intervals",
     ylab = "Average Number of Steps",
     ylim = c(0, max(mean_steps_weekday)))

plot(mean_steps_weekend, 
     type = "l", 
     main = "Time Series Plot of the average steps for each time-interval: Weekdends",
     cex.main = .9,
     xlab = "5-Minute Intervals",
     ylab = "Average Number of Steps",
     ylim = c(0, max(mean_steps_weekday)))
# Reset
par(mfrow = c(1,1))
```